\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{hyperref}

\setstretch{1.1}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.8em}

\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}

\begin{document}

\begin{center}
    {\LARGE \textbf{UMass Lowell Summer 2026 KCS Research Proposal}} 
    
    \vspace{0.5em}
    
    {\large Jegannarayanan Sivakasiulaganathan, Student ID: 02197006}
    
    \vspace{0.25em}
    {\large Advisor: Dr. Paul Downen}
    
    \vspace{0.25em}

    {\large Graduate Student, Computer Science}
    
    \vspace{0.25em}
    
    {\large Expected Graduation: May 2027}
\end{center}

\vspace{1.5em}
\section*{Summary}
This Research Grant Proposal will focus on the implementation and proofs of
novel, cutting edge Dependent Intersection type programming structures in Rocq.



Dependent Types allow programmers to express program invariants in their types.
Traditional type systems allow partial correctness, but dependent types 
enforce invariants which are useful for preventing bad program behavior.
One classic case is a list dependent type which encodes how many elements it has - 
if a type system correctly encodes this,
the programmer cannot mistakenly refer to an empty list as if it had elements.

Dependent Intersection Types allow for intersection types where
we have terms with type $A \cap B$, and $B$ depends on $A$.
Intuitively, this lets us say: “this value has property A, and also property B that is defined using A.”


% TODO: Remove the below if word constrained.

Dependent Intersection types are not available natively in Rocq. As a result, 
programmers must simulate them using existing features which introduces 
additional proof overhead and complexity

Right now, you can simulate dependent intersections, but doing so makes proofs 
significantly harder to construct and maintain. When properties depend on values,
developers—especially programming languages researchers and verification 
engineers—must manually carry equalities, rewrite terms, and introduce extra 
helper lemmas just to convince Rocq that related pieces are consistent. In 
larger projects, like proving a compiler correct or formalizing a type system, 
this turns into a lot of low-value “bookkeeping” work that slows progress and 
makes proofs harder to understand and maintain.

Dependent intersection types directly reduce this burden by letting the type 
system track these relationships automatically. Instead of constantly proving 
that different pieces line up, the system already knows they do. This can 
greatly shorten proofs, remove repetitive steps, and make large formal 
developments more manageable.

Just as importantly, this work has implications for the core design of the 
proof assistant itself. Today, many different features and patterns are needed 
to simulate what dependent intersections could express more directly. By 
introducing a single, well-designed primitive, we can replace several of these 
workarounds. This leads to a smaller and simpler kernel—the trusted core of 
the system—which is easier to audit and reason about for correctness.

At the same time, even though the kernel becomes smaller, it becomes more 
capable (able to describe more kinds of relationships directly). This is 
valuable because it means we can say more with less machinery: fewer special 
cases, fewer encodings, and a cleaner foundation overall. For researchers and 
advanced users, this is a major benefit—it improves both the usability of the 
system and confidence in its correctness, while enabling more ambitious 
verification efforts.

\section*{Planned Work}
The planned work is first to port new dependent type structures that Dr. Downen and Shriya Thakur are working on
to Rocq. This involves implementing the proposed types in the proof assistant Rocq.

Next, what is planned is to prove certain properties about these new types which make them desirable to be used.
Some of these are: 

\begin{enumerate}
  \item soundness - The implementation guarantees that if type checking succeeds, the execution will 
    not yield type related runtime errors.
  \item adequacy - The implementation guarantees that the intended meaning of a type matches the actual
    behavior of the system.
\end{enumerate}

Finally, a compiler written in Haskell is planned to allow programmers to write and execute
programs using the new type system.
This involves implementing a tokenizer, parser, and type checker.

A good estimate for work to be done is implementation in the 1st month, proofs in the 2nd and 3rd months,
and a compiler for evaluation as a stretch goal.

\section*{Expected Results}

The expected results from this research effort are an implementation of aforementioned dependent type structures,
as well as proofs of their soundness and adequacy.

Lastly, a compiler to evaluate it, is a stretch goal. This compiler would be able to type check and reason about
the newly proposed types, and evaluate/run code.

\end{document}

I think the thing that is really missing is some more details about the specifics in the system and the outcomes, that really makes this cutting-edge *research* instead of a general project. For example, what kind of dependent type system is it and what makes it different from the other ones (in this case, the dependent intersection types)? Why are those exciting and an improvement from the current systems (in this case, the ability to encode and express many advanced type system features with a smaller kernel language)? What is the outcome that makes this summer project exciting to do, and why this is a necessary/useful followup step from the starting point that Shriya and I are working on?
